### 2014-06-03 writing/S'14 paper/abstract.tex
Revisions focused on the join table section

\begin{abstract}
Supervised recursive neural network models (RNNs) for sentence meaning
have been successful in a wide array of sophisticated language tasks,
but it remains an open question whether they can achieve the same
results as compositional semantic grammars based in logical forms. We
address this question directly by using a logical grammar to generate
controlled data sets encoding the relationships (entailment, synonymy,
contradiction) between pairs of expressions and evaluating whether
each of two classes of neural model --- plain RNNs and recursive
neural tensor networks (RNTNs) --- can learn those relationships
correctly. Our first experiment confirms that both models can learn
the basic algebra of logical relations involved. Our second and third
experiments expand on this result to cover complex recursive
structures and sentences involving quantification. We find that the
plain RNN achieves only mixed results on the second and third
experiments, but that the stronger RNTN models generalized well in
every case.
\end{abstract}

>>>>

\begin{abstract}
  Supervised recursive neural network models (RNNs) for sentence
  meaning have been successful in a wide array of sophisticated
  language tasks, but it remains an open question whether they can
  achieve the same results as compositional semantic grammars based in
  logical forms. We address this question directly by using a logical
  grammar to generate controlled data sets encoding the relationships
  (e.g., entailment, contradiction) between pairs of expressions and
  evaluating whether each of two classes of neural model --- plain
  RNNs and recursive neural tensor networks (RNTNs) --- can learn
  those relationships correctly. Our first experiment evaluates
  whether these models can learn the basic algebra of logical
  relations involved. Our second and third experiments extend this
  evaluation to complex recursive structures and sentences involving
  quantification.  We find that the plain RNN achieves only mixed
  results on all three experiments, whereas the stronger RNTN models
  generalizes well in every setting.
\end{abstract}

### 2014-06-03 writing/S'14 paper/intro.tex
Revisions focused on the join table section

We address this question in the context of \ii{natural language
  inference} (also known as \ii{recognizing textual entailment};
\cite{dagan2006pascal}), in which the goal is to determine the core
inferential relationship between two sentences. Much of the
theoretical work on this task (and some successful implemented models
\cite{maccartney2009natural,watanabe2012latent}) involves \ii{natural
  logics}, which are formal systems that define sound rules of
inference between words, phrases, and sentences of language without
the need of intermediate representations in an artificial logical
language. Following \cite{bowman2013can}, we use the natural logic
developed by \cite{maccartney2009extended} as our formal model. This
logic defines seven core relations of synonymy, entailment,
contradiction, and mutual consistency, and it provides rules of
semantic combination for projecting these relations from the lexicon
up to complex phrases. The formal properties and inferential strength
of this system are now well-understood
\cite{Icard:Moss:2013,Icard:Moss:2013:LILT}.

>>>>

We address this question in the context of \ii{natural language
  inference} (also known as \ii{recognizing textual entailment};
\cite{dagan2006pascal}), in which the goal is to determine the core
inferential relationship between two sentences. Much of the
theoretical work on this task (and some successful implemented models
\cite{maccartney2009natural,watanabe2012latent}) involves \ii{natural
  logics}, which are formal systems that define rules of inference
between natural language words, phrases, and sentences without the
need of intermediate representations in an artificial logical
language. Following \cite{bowman2013can}, we use the natural logic
developed by \cite{maccartney2009extended} as our formal model. This
logic defines seven core relations of synonymy, entailment,
contradiction, and mutual consistency, as summarized in
Table~\ref{b-table}, and it provides rules of semantic combination for
projecting these relations from the lexicon up to complex phrases. The
formal properties and inferential strength of this system are now
well-understood \cite{Icard:Moss:2013,Icard:Moss:2013:LILT}.

### 2014-06-03 writing/S'14 paper/join.tex
Revisions focused on the join table section

In order for a model to be able to learn to mimic the behavior of a relational logic like the one presented here from a finite amount data, it must be able to learn to deduce new relations from seen relations. The simplest such deductions involve facts about the relations themselves that do not involve considering the internal structure of the things being compared. For example, given that $a\sqsupset b$ and $b\sqsupset c$ one can conclude that $a\sqsupset c$ by the transitivity of $\sqsupset$, even without understanding $a$, $b$, or $c$. These seven relations support more than just transitivity: MacCartney and Manning's \cite{maccartney2009extended} join table defines 32 valid inferences that can be made on the basis of pairs of relations of the form $a R b$ and $b R' c$, including several less intuitive ones such as that if $a \natneg b$ and $b~|~c$ then $a \sqsupset c$.

\begin{figure}[t]
\begin{center}
\begin{tabular}{ll|lll}
$a := \{1, 3, 4, 5, 6\}	$&~~~&~~~& $a \equiv a$	    	& $a~\#~c$	\\
$b := \{0, 3, 5, 6\}	$&~~~&~~~& $e \sqsupset c$	&$b \smallsmile c$		\\
$c := \{1, 2, 3, 4\}	$&~~~&~~~& $d \smallsmile e$	& $b~\#~d$		\\
$d := \{0, 4\}		$&~~~&~~~& $a \sqsubset e$	& $ d \equiv d$	\\
$e := \{1, 2, 3, 4, 5, 6\}$&~~~&~~~& $a~\#~b$		& ... 	\\
\end{tabular}
\end{center}

\caption{Some sample randomly generated sets, and some of the relations defined between them.  \label{lattice-figure}} 
\end{figure}

I test the model's ability to learn this behavior by creating an artificial dataset of terms which represent sets of numbers. Since MacCartney and Manning's set of relations hold between sets as well as between sentences, I can use the underlying set structure to generate the all of the relations that hold between any pair of these terms, as in Figure \ref{lattice-figure}. I train the model defined above on a subset of these relations, but rather then presenting the model with a pair of tree-structured sentences as inputs, simply present it with two single terms, each of which corresponds to a single vector in the (randomly initialized) vocabulary matrix $V$, ensuring that the model has no information about the terms being compared except the relations between them.

I generate 80 randomly generated sets drawing from the same seven elements, and create a dataset consisting of the relations between every pair of sets, yielding 6400 pairs. 3200 of these pairs were then chosen as a test dataset, and that test dataset was further split into the 2960 examples that can be provably derived from the test data using MacCartney and Manning's join table (or by the symmetry of the relations in about half of the cases) and the 240 that cannot. % TODO: Say more about symmetry?

We found that the RNTN model worked best with 11 dimensional vector representations for the 80 sets and a 90 dimensional feature vector for the classifier. This model was able to correctly label 99.3\% of the derivable test examples, and 99.1\% of the remaining examples. The simpler RNN model worked best with 11 and 75 dimensions, respectively, but was able to achieve accuracies of only 90.0\% and 87.\%, respectively.

Discussion to be filled in based on Monday's results.\\...\\...\\...\\...\\...\\...

% Points for discussion: 
% - Fairly simple interpretation: RNTN generalized, RNN didn't
% - Future work: Can this generalize to models with huge numbers of entities, or does this result depend on the examples reflecting a small model in any way? Can't exhaustively generate large models without getting \#s - need sampling.

>>>>

If a model is to learn the behavior of a relational logic like the one
presented here from a finite amount data, it must be able to learn to
deduce new relations from seen relations in a sound and complete
manner. The simplest such deductions involve atomic statements using
the relations in Table~\ref{b-table}. For instance, given that $a
\natrev b$ and $b \natrev c$, one can conclude that $a \natrev c$, by
basic set-theoretic reasoning (transitivity of $\natrev$). Similarly,
from $a \natfor b$ and $b \natneg c$, it follows that $a \natalt c$.
The full set of sound inferences of this form is summarized in
Table~\ref{tab:jointable}.

% about the relations themselves that do not depend on the
% internal structure of the things being compared. For example, given
% that $a\sqsupset b$ and $b\sqsupset c$ one can conclude that
% $a\sqsupset c$ by the transitivity of $\sqsupset$, even without
% understanding $a$, $b$, or $c$. These seven relations support more
% than just transitivity: MacCartney and Manning's
% \cite{maccartney2009extended} join table defines 32 valid inferences
% that can be made on the basis of pairs of relations of the form $a R
% b$ and $b R' c$, including several less intuitive ones such as that if
% $a \natneg b$ and $b~|~c$ then $a \sqsupset c$.

\begin{table}[htp]
  \centering  
  \setlength{\arraycolsep}{12pt}
  \renewcommand{\arraystretch}{1.1}
  %\newcommand{\rcg}{\cellcolor[gray]{.9}}
  \newcommand{\rcg}{}
  $\begin{array}[t]{c@{ \ }|*{7}{c}|}
    %\hline
    \multicolumn{1}{c}{}
             & \nateq     & \natfor     & \natrev     & \natneg    & \natalt     & \natcov     & \multicolumn{1}{c}{\natind} \\
    \cline{2-8}
    \nateq   & \rcg\nateq & \natfor     & \natrev     & \natneg    & \natalt     & \natcov     & \natind\\
    \natfor  & \natfor    & \rcg\natfor & \natind     & \natalt    & \natalt     & \natind     & \natind\\
    \natrev  & \natrev    & \natind     & \rcg\natrev & \natcov    & \natind     & \natcov     & \natind\\
    \natneg  & \natneg    & \natcov     & \natalt     & \rcg\nateq & \natrev     & \natfor     & \natind\\
    \natalt  & \natalt    & \natind     & \natalt     & \natfor    & \rcg\natind & \natfor     & \natind\\
    \natcov  & \natcov    & \natcov     & \natind     & \natrev    & \natrev     & \rcg\natind & \natind\\
    \natind  & \natind    & \natind     & \natind     & \natind    & \natind     & \natind     & \rcg\natind \\
    \cline{2-8}
  \end{array}$
  \caption{Full set of inferences between natural logic relations.
    Given $a \mathbin{R} b$ and $b \mathbin{S} c$, where $R$ and $S$
    are the row and column relations, respectively, and $a$, $b$, and
    $c$ are arbitrary formulae, the table provides the relation
    $a \mathbin{T} c$.}
  \label{tab:jointable}
\end{table}

To test our models' ability to learn this relational structure, we
create small models for our logic in which terms denote sets of
entities from a single domain of seven entities (integers).
Figure~\ref{lattice-figure} depicts a small model of this form. The
lattice structure gives the full model, in which all the statements on
the right can be proved. The highlighted nodes represent a
hypothetical training set drawn from this larger space, which allows
us to prove only a subset of the full set of true statements.

\begin{figure}[htp]
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \newcommand{\labelednode}[4]{\put(#1,#2){\oval(1.5,1)}\put(#1,#2){\makebox(0,0){$\begin{array}{c}#3\\\{#4\}\end{array}$}}}
    \setlength{\unitlength}{1cm}
    \begin{picture}(5,5.5)
      \labelednode{2.50}{5}{a}{1,2,3}

      \put(0.75,4){\line(3,1){1.5}}
      \put(2.5,4){\line(0,1){0.5}}
      \put(4.25,4){\line(-3,1){1.5}}

      \labelednode{0.75}{3.5}{b}{1,2}
      \labelednode{2.50}{3.5}{c}{1,3}
      \labelednode{4.25}{3.5}{d}{2,3}

      \put(0.75,2.5){\line(0,1){0.5}}
      \put(0.75,2.5){\line(3,1){1.5}}

      \put(2.5,2.5){\line(-3,1){1.5}}
      \put(2.5,2.5){\line(3,1){1.5}}

      \put(4.25,2.5){\line(0,1){0.5}}
      \put(4.25,2.5){\line(-3,1){1.5}}

      \labelednode{0.75}{2}{e}{1}
      \labelednode{2.50}{2}{f}{2}
      \labelednode{4.25}{2}{g}{3}

      \put(2.5,1){\line(-3,1){1.5}}
      \put(2.5,1){\line(0,1){0.5}}
      \put(2.5,1){\line(3,1){1.5}}

      \labelednode{2.5}{0.5}{h}{}
    \end{picture}
    \caption{Simple boolean model.}
  \end{subfigure}
  \qquad
  \begin{subfigure}[t]{0.5\textwidth}
    \centering
    \setlength{\tabcolsep}{12pt}
    \begin{tabular}[b]{c  c}
      \toprule
      Train & Test \\
      \midrule
                    & $b \nateq b$ \\
      $b \natcov c$ &               \\
                    & $b \natcov d$ \\
                    & $b \natrev e$ \\
      $c \natcov d$ &               \\
      $c \natrev e$ &               \\
                    & \strikeout{$c \natneg f$} \\
      $c \natrev g$ &               \\ 
                    & $e \natfor b$ \\
      $e \natfor c$ &               \\[-1ex]
      $\vdots$      & $\vdots$ \\
      \bottomrule
    \end{tabular}

    \caption{An example train/test split from the full set
      of statements one can make about the model.
      Statements not provable from the test data are crossed out.}
  \end{subfigure}  
  % \begin{tabular}{ll|lll}
  %   $a := \{1, 3, 4, 5, 6\}	$&~~~&~~~& $a \equiv a$	    	& $a~\#~c$	\\
  %   $b := \{0, 3, 5, 6\}	$&~~~&~~~& $e \sqsupset c$	&$b \smallsmile c$		\\
  %   $c := \{1, 2, 3, 4\}	$&~~~&~~~& $d \smallsmile e$	& $b~\#~d$		\\
  %   $d := \{0, 4\}		$&~~~&~~~& $a \sqsubset e$	& $ d \equiv d$	\\
  %   $e := \{1, 2, 3, 4, 5, 6\}$&~~~&~~~& $a~\#~b$		& ... 	\\
  % \end{tabular}
  \caption{Some sample randomly generated sets, and some of the relations defined between them.}
  \label{lattice-figure}
\end{figure}

% We test the model's ability to learn this behavior by creating
% artificial data sets of terms which represent sets of numbers. Since
% MacCartney and Manning's set of relations hold between sets as well as
% between sentences, we can use the underlying set structure to generate
% the all of the relations that hold between any pair of these terms, as
% in Figure \ref{lattice-figure}. We train the model defined above on a
% subset of these relations, but rather then presenting the model with a
% pair of tree-structured sentences as inputs, simply present it with
% two single terms, each of which corresponds to a single vector in the
% (randomly initialized) vocabulary matrix $V$, ensuring that the model
% has no information about the terms being compared except the relations
% between them.

In our experiments, we create 80 randomly generated sets drawing from
the same domain of seven element. This yields a data set consisting of
6400 statements about pairs of entities. 3200 of these pairs are
chosen as a test set, and that test set is further reduced to the 2960
examples that can be provably derived from the test data using
Table~\ref{tab:jointable}.

% TODO: Say more about symmetry?
We trained versions of both the RNN model and the RNTN model on these
data sets. In both cases, the models were implemented exactly as
described in Section~\ref{methods}, but since the items being compared
are single terms rather than full tree structures, the composition
layer was not involved, and the two models differed only in which
layer function was used for the comparison layer. We simply present
the models with two single terms, each of which corresponds to a
single vector in the (randomly initialized) vocabulary matrix $V$,
ensuring that the model has no information about the terms being
compared except the relations between them. We found that the RNTN
model worked best with 11-dimensional vector representations for the
80 sets and a 90-dimensional feature vector for the classifier. This
model was able to correctly label 99.3\% of the derivable test
examples, and 99.1\% of the remaining examples. The simpler RNN model
worked best with 11 and 75 dimensions, respectively, but was able to
achieve accuracies of only 90.0\% and 87.\%, respectively.

% and create a dataset consisting of the relations between every pair of
% sets, yielding 6400 pairs. 3200 of these pairs are then chosen as a
% test dataset, and that test dataset was further split into the 2960
% examples that can be provably derived from the test data using
% MacCartney and Manning's join table (or by the symmetry of the
% relations in about half of the cases) and the 240 that
% cannot. % TODO: Say more about symmetry?

% We tested a version of both the RNN model and the RNTN model on these
% data. In both cases, the models were implemented exactly as described
% in Section~\ref{methods}, but since the items being compared are
% single terms rather than full tree structures, the composition layer
% was not used, and the two models differed only in which layer function
% was used for the comparison layer. We found that the RNTN model worked
% best with 11 dimensional vector representations for the 80 sets and a
% 90 dimensional feature vector for the classifier. This model was able
% to correctly label 99.3\% of the derivable test examples, and 99.1\%
% of the remaining examples. The simpler RNN model worked best with 11
% and 75 dimensions, respectively, but was able to achieve accuracies of
% only 90.0\% and 87.\%, respectively.

% TODO: T-sne the RNTN model's embeddings.

These results are fairly straightforward to interpret. The RNTN model
was able to accurately encode the relations between the terms in the
geometric relations between their vectors, and was able to then use
that information to recover relations that were not overtly included
in the training data. In contrast, the RNN model was able to
approximate this behavior only incompletely. It is possible but not
likely that it could be made to find a good solution with further
optimization on different learning algorithms, or that it would do
better on a larger universe of sets for which there would be a larger
set of training data to learn from, but the RNTN is able to achieve
these effects in the well-studied framework of Section~\ref{methods}.

### 2014-06-03 writing/S'14 paper/methods.tex
Revisions focused on the join table section

The models that we study are are centered aronund a recursively applied composition function which is meant to mimic the recursive construction of meanings in formal semantics. In this scheme, pairs of words are merged into phrase representations by a function that maps from representations of length $2N$ to representations of length $N$. These phrase representations are then further merged with other words and phrases until the entire phrase or sentence being evaluated is represented in a single vector. This vector is then used as the input to a classifier and used in a supervised learning task.

We use the model architecture proposed in \citet{bowman2013can} and depicted in Figure \ref{sample-figure}. The two phrases being compared are built up separately on each side of the tree using the same composition function until they have each been reduced to single vectors. Then, the two phrase vectors are fed into a separate comparison layer that is meant to generate a feature vector capturing the relation between the two phrases. The output of this layer is then given to a softmax classifier, which in turn produces a hypothesized distribution over the seven relations.

For a composition layer, we evaluate models with both the plain neural network layer function (eq. \ref{rnn}) and the RNTN layer function proposed in \citet{chen2013learning} (eq. \ref{rntn}). A sigmoid nonlinearity (elementwise $\tanh$) is applied to the output of either layer function, following Socher et al.
\begin{gather} \label{rnn}
y_{RNN} = f(\mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + b_i)\\ % TODO: Add column vectors?
\label{rntn}
y_{RNTN} = f(\vec{x}^{(l)T} \mathbf{A}^{[1...N]} \vec{x}^{(r)} + \mathbf{B} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{c})
\end{gather} % TODO: Explain third order tensor parameter?

The comparison layer uses the same type function template as the composition function (either an NN layer or an NTN layer) with independantly learned parameters and a separate nonlinearity function. Rather than use a $\tanh$ nonlinearity here, we found a substantial improvement in performance by using a rectified linear function for $f_{b}$. In particular, we use the leaky rectified linear function \cite{maasrectifier}: $f_{b}(\vec{x})=\max(\vec{x}, 0) + 0.01\min(\vec{x}, 0)$,  applied elementwise.

@@

We train the model with minibatch stochastic gradient descent (SGD) with learning rates computed using AdaGrad \cite{duchi2011adaptive} from a starting rate of 0.2. The parameters (including the vocabulary) are initialized randomly using a uniform distribution over $[-0.1, 0.1]$. % TODO: Update

\begin{figure}
\begin{center}
\includegraphics[scale=0.35]{model.eps}
\end{center}
\caption{The model structure used to compare \ii{no dog} and \ii{(not all) dog}. The same structure is used for both the RNN and RNTN layer functions. \label{sample-figure}} 
\end{figure}

\ii{Source code and generated data will be released after the conclusion of the review period.} % TODO: Or upon request? Attach anonymized code?

>>>>

We study neural models that adhere to the linguistic \ii{principle of
  compositionality}, which says that the meanings for complex
expressions are derived from the meanings of their constituent parts
via specific combination functions \cite{Partee84,Janssen97}. In our
distributional model, word meanings are vectors of length $N$. The
combination function concatenates pairs of them to form vectors of
length $2N$ and maps those larger vectors back into length $N$
vectors, which can then be merged again to create more complex
phrases. Once the entire sentence-level representation has been
derived, it fed to a classifier trained in a supervised manner.

% The models that we study are centered around a recursively applied composition function which is meant to mimic the recursive construction of meanings in formal semantics. In this scheme, pairs of words are merged into phrase representations by a function that maps from representations of length $2N$ to representations of length $N$. These phrase representations are then further merged with other words and phrases until the entire phrase or sentence being evaluated is represented in a single vector. This vector is then used as the input to a classifier and used in a supervised learning task.

We use the model architecture proposed in \cite{bowman2013can} and
depicted in Figure~\ref{sample-figure}. The two phrases being compared
are built up separately on each side of the tree, using the same
composition function, until they have each been reduced to single
vectors. The resulting vectors are fed into a separate comparison
layer that is meant to generate a feature vector capturing the
relation between the two phrases. The output of this layer is then
given to a softmax classifier, which in turn produces a hypothesized
distribution over the seven relations represented in Table~\ref{b-table}.

For a composition layer, we evaluate models with both the plain neural
network layer function (eq.~\ref{rnn}) and the RNTN layer function
proposed in \citet{chen2013learning} (eq.~\ref{rntn}). A sigmoid
nonlinearity (element-wise $\tanh$) is applied to the output of either
layer function, following \cite{socher2013acl1}.
%
\begin{gather} \label{rnn}
y_{\textit{RNN}} = f(\mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + b_i)\\ % TODO: Add column vectors?
\label{rntn}
y_{\textit{RNTN}} = f(\vec{x}^{(l)T} \mathbf{A}^{[1 \ldots N]} \vec{x}^{(r)} + \mathbf{B} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{c})
\end{gather} % TODO: Explain third order tensor parameter?
%
Here, the $l$ and $r$ superscripts indicate the vector representations
for the left and right children of the local structure interpreted as
$y$. The RNN concatenates them, multiplies them by an $N \times 2N$
matrix of learned weights, and applies the element-wise non-linearity
to the resulting vector (plus a bias term). The RNTN has the same
basic structure, but with the addition of a tensor $\mathbf{A}$,
dimension $2N \times N \times N$, modeling interactions between the
child vectors.

The comparison layer uses the same type function template as the
composition function (either an NN layer or an NTN layer) with
independantly learned parameters and a separate nonlinearity function.
Rather than use a $\tanh$ nonlinearity here, we found a substantial
improvement in performance by using a rectified linear function for
$f_{b}$. In particular, we use the leaky rectified linear function
\cite{maasrectifier}: $f_{b}(\vec{x})=\max(\vec{x}, 0) +
0.01\min(\vec{x}, 0)$, applied element-wise.

@@

We train the model with minibatch stochastic gradient descent (SGD)
with learning rates computed using AdaGrad \cite{duchi2011adaptive}
from a starting rate of 0.2. The parameters (including the vocabulary)
are initialized randomly using a uniform distribution over $[-0.1,
0.1]$. % TODO: Update
Source code and generated data will be released after the conclusion
of the review period.

\begin{figure}[htp]
  \centering
  \includegraphics[scale=0.35]{model.eps}
  \caption{The model structure used to compare \ii{no dog} and \ii{(not all) dog}. 
    The same structure is used for both the RNN and RNTN layer functions.} 
  \label{sample-figure}
\end{figure}

%\ii{Source code and generated data will be released after the conclusion of the review period.} % TODO: Or upon request? Attach anonymized code?
